---
title: "relatedness"
author: "Rafa Rico"
date: "2025-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Modified code from https://hackmd.io/@MEC-lab-team/r1XvULhVxx

Libraries

```{r}
library(here)
library(dplyr)
library(tidyverse)
```

Data

```{r}
here("/ngsrelate")
chr1 <- read.table("relatedness_chr1", header = T)
chr2 <- read.table("relatedness_chr2", header = T)
pop <- read_csv("../gen_samples.csv")
```

Using the relatedness metric rab - (Hedrick & Lacy's relatedness estimator)

```{r}
ngs_matrix <- chr1 %>% select(ida, idb, rab)
```

Convert to wide form: create a symmetrical matrix with individuals as rows and columns.

```{r}
# Create a symmetrical matrix

relatedness_matrix <- ngs_matrix %>%
  pivot_wider(names_from = idb, values_from = rab) %>%
  column_to_rownames("ida")

# Ensure symmetry
relatedness_matrix[lower.tri(relatedness_matrix)] <- t(relatedness_matrix)[lower.tri(relatedness_matrix)]

```

#Heatmap

```{r}
library(pheatmap)

library(RColorBrewer)
#display.brewer.all(colorblindFriendly = T)
heat_colors <- brewer.pal(6, "OrRd")

pheatmap(relatedness_matrix, color=heat_colors,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         display_numbers = FALSE,
         main = "Pairwise Relatedness (rab)")
```

#Stacked bar Plots

Add population info for ida and idb

```{r}
chr1 <- chr1 %>% left_join(pop,by=c("ida" = "sample_name")) %>% rename(PopA = population)

chr1 <- chr1 %>% left_join(pop,by=c("idb" = "sample_name")) %>% rename(PopB = population)
```

Create combined population pair (alphabetical order to avoid duplicates) 

```{r}
chr1 <- chr1 %>%
  mutate(PopPair = ifelse(PopA < PopB,
                          paste(PopA, PopB, sep = "."),
                          paste(PopB, PopA, sep = ".")))
```

```{r}
pops_chr1 <- chr1 %>% filter(PopA==PopB)
```


## Within-pop comparisons 

Filter to within-population comparisons

```{r}
within_df <- chr1 %>% filter(PopA == PopB)%>%
  select(ida, idb, )
```

Stacked bar plot

```{r}
# Define bin levels and breaks
rab_levels <- c("0-0.1", "0.1-0.2", "0.2-0.3", "0.3-0.4", 
                "0.4-0.5", "0.5-0.6", "0.6-1")
rab_breaks <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1)

# Bin rab values
within_df$rab_bin <- cut(within_df$rab,
                         breaks = rab_breaks,
                         labels = rab_levels,
                         include.lowest = TRUE, # This ensures that the very first bin includes the lowest value
                         right = TRUE) #right=TRUE is to specify that the upper bound is included (i.e. 0.01 is included in the interval [0,0.01])

# Summarize counts per population and bin
stacked_df <- within_df %>%
  group_by(PopPair, rab_bin) %>% # Group data by both Pop pair and rab_bin
  summarise(count = n(), .groups = "drop") %>% # Count how many pairwise comparisons falls into that bin for that PopPair. .groups = "drop" tells dplyr to remove all grouping after this step cuz the next function 'complete' expects an ungrouped data frame 
  complete(PopPair, rab_bin = rab_levels, fill = list(count = 0)) %>% #ensures all combinations of PopPair and rab_bin exist in the dataset, even if they were missing. So if there are no rab values between 0.04 and 0.08 for a PopPair, it will set the count to 0. 
  group_by(PopPair) %>% #Regroup by PopPair
  mutate(proportion = count / sum(count)) %>%
  ungroup() #Removes grouping structure again, leaving a clean, regular data.frame that’s ready for plotting 

```


plotting

```{r}
p <- ggplot(stacked_df, aes(x = factor(PopPair), y = proportion, fill = factor(rab_bin))) +
  geom_bar(stat = "identity",width=0.5) + coord_flip() +
  ylab("Proportion of pairwise comparisons") +
  xlab("Population") +
  scale_fill_grey(name = "Pairwise relatedness") +
  scale_y_continuous(labels = c(
    "0.00" = "0",
    "0.25" = "0.25",
    "0.50" = "0.5",
    "0.75" = "0.75",
    "1.00" = "1"),expand = c(0, 0))+
  theme_classic() +
  theme(axis.text=element_text(size=20),
        axis.title=element_text(size=22,face="bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 22,face="bold"))

print(p)
```


#Heatmap with mean rab for each PopPair

Calculate mean and std dev for each PopPair

```{r}
heat_df <- chr1 %>%
  group_by(PopPair) %>%
  summarise(
    mean_rab = mean(rab, na.rm = TRUE),
    sd_rab = sd(rab, na.rm = TRUE),.groups = "drop")

```

Split PopPair into two columns for heatmap

```{r}
heat_df <- heat_df %>%
  separate(PopPair, into = c("PopA", "PopB"), sep = "\\.")

# Make it symmetric
symmetric_df <- bind_rows(
  heat_df,
  heat_df %>% rename(PopA = PopB, PopB = PopA)
)
```

Calculate mask after factor conversion

```{r}
population_order <- unique(symmetric_df$PopA)
symmetric_df$PopA <- factor(symmetric_df$PopA, levels = rev(population_order))
symmetric_df$PopB <- factor(symmetric_df$PopB, levels = rev(population_order))
```


```{r}
symmetric_df <- symmetric_df %>% distinct() %>%
  mutate(
    mask = as.integer(PopA) > as.integer(PopB),  # Lower triangle
    mean_rab_masked = ifelse(mask, NA, mean_rab) # Set to NA if in lower triangle
  )
```

Plot
```{r}
library(RColorBrewer)
heat_colors <- brewer.pal(6, "OrRd")
```

```{r}
ggplot(symmetric_df, aes(x = PopA, y = PopB, fill = mean_rab_masked)) +
  geom_tile(color = "white") +
  geom_tile(data = subset(symmetric_df, is.na(mean_rab_masked)), fill = "grey", color = "white") +  # Black fill for masked cells
  #geom_text(aes(label = ifelse(!is.na(mean_rab_masked), sprintf("%.6f", mean_rab_masked), "")), size = 6) +
  scale_fill_gradientn(colors = heat_colors, name = "Mean pairwise relatedness",labels = function(x) sprintf("%.6f", x),
                       na.value = "black") +  # Also ensures NA fill is black
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank(),
    legend.key.height = unit(1, "cm")
  )
```

